---
title: "Provider inference with fit_pci"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Provider inference with fit_pci}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`fit_pci()` provides one interface for provider-level tests, empirical-null
calibration, confidence intervals, and diagnostics. It takes summaries from an
outcome model: estimates and standard errors, observed and expected counts,
individual null probabilities, or precomputed scores. The outcome model and its
covariance calculation remain the user's responsibility.

```{r setup}
library(EmpiNull)
```

## A complete first example

Suppose a fitted model supplies observed and expected event counts for six
providers. An observed/expected ratio of one is the reference. The following
call calculates two-sided Poisson mid-p values and ratio intervals:

```{r first-example}
providers <- data.frame(
  id = paste0("provider_", seq_len(6)),
  observed = c(0, 8, 14, 23, 37, 61),
  expected = c(3, 10, 15, 20, 25, 40)
)
pci <- fit_pci(
  id = providers$id,
  observed = providers$observed,
  expected = providers$expected,
  statistic = "poisson_midp",
  calibration = "theoretical"
)
pci$results[, c("provider_id", "estimate", "p_value", "direction",
                 "ci_lower", "ci_upper", "ci_scale")]
pci$diagnostics$warning_codes
```

`direction` is -1 for a significantly lower result, +1 for a significantly
higher result, and 0 otherwise. It describes the statistical direction, without
assigning clinical desirability. The default significance level is 0.05, and
intervals use confidence level `1 - alpha`. Tests are per provider: no
multiple-testing adjustment is applied automatically.

## Choose the statistic

The `statistic` argument identifies the supplied information and initial test.
Use `alternative = "two.sided"`, `"greater"`, or `"less"` to choose the tail.

| `statistic` | Required input | Initial test |
|:--|:--|:--|
| `"z"` | `z` | Normal score supplied by the user |
| `"p_value"` | `p_value`, `direction` | Supplied two-sided p-value converted to a signed normal score |
| `"wald"` | `estimate`, `se` | Normal or Student-t test, optionally on the log/logit scale |
| `"log_ratio"` | `estimate`, `se` | Log-ratio Wald test; SE is on the log scale by default |
| `"poisson_midp"` | `observed`, `expected` | Poisson half-mass tails |
| `"poisson_exact"` | `observed`, `expected` | Inclusive Poisson tails with the convention specified below |
| `"poisson_score"` | `observed`, `expected` | Pearson score with a normal reference |
| `"poibin"` | `observed`, `probabilities` | Distribution of a sum of independent Bernoulli variables |
| `"binary_score"` | `observed`, `probabilities` | Standardized Bernoulli count with a normal reference |
| `"resampling"` | `observed`, `probabilities` or `eta` | Monte Carlo Bernoulli-count tails, optionally integrating normal posterior effects |

`"poibin_exact"` is an alias for `"poibin"`. Deterministic binary methods also
accept `eta` and `re_mean`; these imply probabilities
`plogis(reference + eta + re_mean)`, and require zero `re_var`.

Provide one entry per provider for the primary input. Standard errors,
references, degrees of freedom, and specified null parameters can be scalars
or provider-length vectors. Observation-level input uses a list with one
numeric vector per provider. For a single provider, a numeric vector is also
accepted. `id` must contain unique nonmissing identifiers.

## Choose the null calibration

For estimated or fixed normal nulls, the function computes

\[
 Z_i^{\mathrm{adj}}=\frac{Z_i-\mu_i}{\sigma_i},\qquad
 p_i=2\Phi(-|Z_i^{\mathrm{adj}}|)
\]

for a two-sided test. One-sided tests use the corresponding normal tail.

| `calibration` | Meaning | Additional information |
|:--|:--|:--|
| `"theoretical"` | Preserve the initial test's tail probabilities | Default; includes t and discrete tails |
| `"fixed"` | Use specified score means and SDs | `null_mean`, `null_sd` |
| `"overall"` | Fit one common null distribution | Robust by default; likelihood via `control$estimator` |
| `"groupwise"` | Fit robust means and SDs within groups | `size` and grouping options, or explicit `groups` |
| `"individualized"` | Gaussian working null with variance `1 + size * phi` | Positive `size`; mean zero or fitted common intercept |
| `"correlated"` | Gaussian working null with mean `sqrt(size) * X %*% zeta` and variance `1 + size * phi` | Positive `size`, optional `covariates` |
| `"cre"` | Mean-zero Gaussian working null with variance `size * phi` | Positive `size` |
| `"moment"` | Gaussian working null matched to Poisson-lognormal moments | Pearson scores, expected counts, and `second_moment`; optional `covariates` |

The structured models use the estimators described in
`?fit_empirical_null` and the package's empirical-null tutorial. An intercept
is included by default for correlated and moment models. Choose the model
according to its assumptions; a shared interface does not make all
calibrations appropriate for every statistic.

### Specified score SD versus an outcome benchmark

`null_sd` is the SD of the provider-score distribution. `reference` is the
outcome value being tested. They answer different questions and can be used
together.

```{r fixed-and-reference}
# A specified score distribution with mean zero and SD 1.81.
fixed <- fit_pci(z = c(-4, 0, 4), calibration = "fixed", null_sd = 1.81)
fixed$results[, c("z_raw", "null_sd", "z_adjusted", "p_value")]

# The corresponding two-sided 5% threshold on the raw-score scale.
1.81 * qnorm(0.975)

# A separate decision: test proportions against a 30% outcome benchmark.
fit_pci(
  estimate = c(0.2, 0.4), se = c(0.04, 0.06),
  statistic = "wald", transform = "logit", reference = 0.3,
  calibration = "fixed", null_sd = 1.81
)$results
```

The value 1.81 is illustrative and is not a package default. It gives a raw
two-sided 5% score cutoff of approximately 3.55. `null_sd = 1.81` does not mean
that the tested rate or ratio equals 1.81.

Theoretical calibration preserves the initial p-values. Fixed calibration
with mean zero and SD one still passes through the score representation, so
it can differ for discrete tails that reach a numerical score floor.

### Overall and groupwise robust fitting

```{r groupwise}
set.seed(23)
provider_size <- seq_len(200)
provider_z <- rnorm(200, mean = 0.15, sd = 1.3)
grouped <- fit_pci(
  z = provider_z, size = provider_size,
  calibration = "groupwise", n_groups = 4,
  grouping = "quantile", psi = "huber"
)
grouped$null_parameters
grouped$diagnostics$warning_codes
```

Robust fitting uses `MASS::rlm()` with an intercept and MAD scale estimate.
`psi = "bisquare"` is the default; `"huber"` is also available. The same
procedure is exposed directly by `empirical_null_groupwise()`.
`calibration = "overall"` fits one robust null using all eligible providers;
`calibration = "groupwise", n_groups = 1` also gives one group. Select the
overall likelihood estimator using `control = list(estimator = "likelihood")`.

Grouping and fit selection are explicit:

- `grouping = "quantile"` uses right-closed size-quantile intervals. Tied
  quantile breaks cause an error; reduce the number of groups or choose a
  different grouping rule.
- `grouping = "rank"` stably sorts providers by size and uses approximately
  equal-count blocks. It can separate tied sizes and can produce fewer than
  `n_groups` nonempty groups for some sample sizes.
- `groups` supplies the group labels directly and takes precedence over
  size-based grouping.
- `fit_mask` selects providers for fitting robust means/scales. All complete
  providers are still tested, including those excluded from fitting.
- `common_mean = TRUE` replaces fitted means with the ordinary pooled mean of
  eligible finite scores. A numeric `common_mean` sets that mean explicitly.
  The previously fitted group scales are retained.

The minimum eligible group size is three by default. Small groups cause an
error unless `control = list(small_group = "theoretical")` requests a recorded
fallback to mean zero and SD one. A nonpositive fitted scale remains an error.
Fitting exclusions and `common_mean` apply only to robust overall/groupwise
modes. Structured likelihood modes require all retained providers.

## Estimate scales and standard errors

With `statistic = "wald"`, `transform` selects the working scale. The input
estimate and reference always remain on the original scale:

| Transformation | Valid estimates/references | Default reference | SE conversion when `se_scale = "original"` |
|:--|:--|:--|:--|
| `"identity"` | Finite real values | 0 | `se` |
| `"log"` | Positive values | 1 | `se / estimate` |
| `"logit"` | Proportions strictly between 0 and 1 | 0.5 | `se / (estimate * (1 - estimate))` |

For example, a 20% rate is entered as `0.2`, not `20`. If the SE is already on
the working scale, use `se_scale = "transformed"`. The `"log_ratio"` shortcut
defaults to the log transform and a transformed SE.

```{r ratio-and-supplied-test}
ratio_fit <- fit_pci(
  estimate = c(0.8, 1.4), se = c(0.15, 0.20),
  statistic = "log_ratio", reference = 1,
  calibration = "fixed", null_mean = 0.2, null_sd = 1.3
)
ratio_fit$results[, c("estimate", "p_value", "ci_lower", "ci_upper")]

# A supplied two-sided model test, with the direction of each estimate.
fit_pci(
  p_value = c(0.01, 0.2), direction = c(-1, 1),
  statistic = "p_value", calibration = "fixed", null_sd = 1.3
)$results[, c("z_raw", "z_adjusted", "p_value")]
```

For a positive ratio with normal Wald inference, a two-sided calibrated
interval is

\[
\exp\{\log(\widehat R)-\mu s\ \mathbin{\pm}
      \Phi^{-1}(1-\alpha/2)\sigma s\},
\]

where \(s\) is the log-scale SE. Use `df` for Student-t Wald inference; finite
t statistics are converted to normal quantiles before calibration, and
intervals invert that transformation. A supplied two-sided p-value plus
direction is assumed to follow a symmetric-tail convention if a one-sided
alternative is requested.

For a zero ratio, `zero_method = "score"` uses `-1 / se`, requiring a
transformed SE and reference one. Its linear upper interval bound is an
approximation recorded as `ZERO_RATIO_SCORE_APPROXIMATION`.
`zero_method = "exclude"` leaves the row present with undefined inference.

## Poisson tails and the two-sided mid-p factor

For provider \(i\), let \(O_i\) be observed events, \(E_i\) be the baseline
expected count, and \(r_0\) be the ratio reference. The null count distribution
is \(X_i\sim\mathrm{Poisson}(E_i r_0)\). Define

\[
L_i=P(X_i<O_i)+\tfrac12 P(X_i=O_i),\qquad
U_i=P(X_i>O_i)+\tfrac12 P(X_i=O_i).
\]

The two-sided mid-p is

\[
p_i^{\mathrm{mid}}=2\min(L_i,U_i).
\]

Since \(L_i+U_i=1\), this lies between zero and one without upper clipping.
If tails are instead written as \(q_i^l=2L_i\) and \(q_i^u=2U_i\), the same
two-sided p-value is \(\min(q_i^l,q_i^u)\), **without division by two**.
Half-mass tails are not guaranteed to give conservative exact test size.

```{r direct-mid-p}
observed <- 13
expected <- 7.385
lower_mid <- ppois(observed - 1, expected) + 0.5 * dpois(observed, expected)
upper_mid <- ppois(observed, expected, lower.tail = FALSE) +
  0.5 * dpois(observed, expected)
direct_p <- 2 * min(lower_mid, upper_mid)
calculated <- fit_pci(
  observed = observed, expected = expected, statistic = "poisson_midp"
)
data.frame(direct_p = direct_p, fit_pci_p = calculated$results$p_value)
```

`"poisson_exact"` uses inclusive tails and a specific two-sided convention:
double the upper tail when `observed > expected * reference`, otherwise double
the lower tail, then cap at 0.999. This is not the probability-ordering
definition used by some other exact-test functions. Under a deterministic
zero-mean null with zero observed events, the two-sided p-value is one.
One-sided `"poisson_midp"` always uses half-mass tails; the `mid_p` argument
does not change the named Poisson methods.

`expected` stays on its baseline scale. Changing `reference` changes the null
mean to `expected * reference`; ratio intervals continue to divide by baseline
`expected`. A zero expected count has no defined observed/expected ratio
interval.

### Pearson scores and the moment model

`"poisson_score"` uses \((O_i-E_i r_0)/\sqrt{E_i r_0}\). For
`calibration = "moment"`, the reference must be one, `size` equals `expected`,
and `second_moment` contains
\(E_i^{(2)}=\sum_j \lambda_{ij}^2\), where
\(E_i=\sum_j\lambda_{ij}\). In general \(E_i^{(2)}\ne E_i^2\).

Alternatively supply Pearson scores as `statistic = "z"` with
`control = list(score_type = "pearson")`, along with `size = E` and
`second_moment = E2`. Mid-p scores are not valid inputs for this moment model.
The model has exact moment expressions but uses a Gaussian working
distribution for tail inference.

## Binary outcomes and resampling

For each provider, `probabilities` contains the individual Bernoulli null
probabilities. `"poibin"` computes their count distribution using dynamic
programming. Values zero and one retain their deterministic meaning.
`"binary_score"` uses

\[
\frac{O_i-\sum_j p_{ij}}
     {\sqrt{\sum_j p_{ij}(1-p_{ij})}}.
\]

```{r binary}
null_probabilities <- list(c(0.1, 0.2, 0.4, 0.6), rep(0.2, 12))
binary <- fit_pci(
  observed = c(2, 4), probabilities = null_probabilities,
  statistic = "poibin"
)
binary$results[, c("p_value", "ci_lower", "ci_upper", "ci_scale")]

simulated <- fit_pci(
  observed = c(2, 4), probabilities = null_probabilities,
  statistic = "resampling", control = list(n_resample = 2000, seed = 7)
)
simulated$results[, c("p_raw", "p_value")]
```

Two-sided Poisson-binomial and resampling tests use half-mass tails by default;
`mid_p = FALSE` uses inclusive tails and clips the doubled smaller tail at one.
One-sided binary tests use inclusive tails unless
`control = list(one_sided_midp = TRUE)`. The `mid_p` switch does not alter the
normal-tail `"binary_score"` test.

Resampling can instead integrate specified normal posterior effects. Supply
`eta`, excluding the effect under test, and optional `re_mean` and `re_var`
lists. For each replicate the method draws independent effects
\(b_{ij}\sim N(m_{ij},v_{ij})\), then independent outcomes with probability
\(\operatorname{logit}^{-1}(\eta_{ij}+r_0+b_{ij})\). Here `reference` is the
additive log-odds effect \(r_0\), default zero.

```{r posterior-resampling}
posterior <- fit_pci(
  observed = c(2, 4), statistic = "resampling",
  eta = lapply(null_probabilities, qlogis),
  re_mean = lapply(null_probabilities, function(p) rep(0, length(p))),
  re_var = lapply(null_probabilities, function(p) rep(0.1, length(p))),
  reference = 0,
  control = list(n_resample = 2000, seed = 7)
)
posterior$results[, c("p_value", "direction")]
```

This sampling model uses one independent effect per observation per replicate;
it does not draw a shared provider effect or a joint posterior covariance.
Increasing `n_resample` reduces simulation variability but changes the realized
tail estimate. Monte Carlo proportions may be zero; no plus-one correction or
simulation-uncertainty interval is applied. The seed is reproducible and the
caller's random-number state is restored.

## Confidence intervals

`ci = "auto"` chooses the following methods. Use `ci = "none"` when only
tests are needed.

| Statistic/calibration | Automatic interval | Reported scale |
|:--|:--|:--|
| Wald or log-ratio | Invert the normal/t working-scale cutoffs and back-transform | Original estimate, ratio, or rate |
| Poisson exact + theoretical | Garwood if expected count is below 100; Byar otherwise | Observed/expected ratio |
| Other Poisson combinations | Invert the selected score with null parameters held fixed | Observed/expected ratio |
| Poisson-binomial or binary score | Invert the count test/score over a common log-odds shift | `log_odds_shift` |
| Supplied Z/p-value or resampling | No automatic original-scale interval | Missing |

For theoretical Poisson inference, `ci = "poisson"` explicitly requests the
classical Garwood/Byar method. If paired with a different Poisson statistic,
the diagnostic `INTERVAL_METHOD_DIFFERS_FROM_TEST` records that choice.
`ci = "invert"` is available for Poisson and deterministic binary methods.

A binary interval varies a common shift \(\delta\) through
`plogis(qlogis(p) + delta)`. It is not automatically an interval for a
standardized rate or observed/expected ratio. With heterogeneous probabilities,
the binary score can be nonmonotone in this shift. A numerical derivative check
withholds a single interval when nonmonotonicity is detected and records
`NONMONOTONE_SCORE_INVERSION`. This is a numerical guard, not a proof of
monotonicity for every input; Poisson-binomial tails provide monotone inversion.

All calibrated intervals hold fitted null parameters fixed. Their estimation
uncertainty is not propagated. One-sided intervals have the appropriate
unbounded or natural-boundary endpoint. A finite root-search limit does not
establish that an endpoint is infinite.

## Controls, returned values, and diagnostics

The `control` list accepts named settings. Defaults and scope are summarized
below; `?fit_pci` documents the full interface.

| Control | Default | Use |
|:--|:--|:--|
| `estimator` | `"robust"` | Overall estimator; groupwise supports robust only |
| `maxit`, `acc` | 1000, `1e-8` | Robust iteration limit and convergence tolerance |
| `tuning` | Huber 1.345; bisquare 4.685 | Robust influence-function tuning |
| `min_group_size`, `small_group` | 3, `"error"` | Minimum fit sample and optional `"theoretical"` fallback |
| `p_grid`, `phi_bounds`, `cutoff` | Model-specific | Structured likelihood estimation; see `?fit_empirical_null` |
| `include_intercept` | `TRUE` | Correlated/moment design matrix |
| `common_intercept` | `FALSE` | Common mean for individualized calibration |
| `p_floor` | `1e-6` | One-sided tail floor for discrete normal-score conversion |
| `seed`, `n_resample` | 1, 10000 | Resampling |
| `one_sided_midp` | `FALSE` | One-sided binary tail convention |
| `normal_threshold` | 100 | Expected-count switch from Garwood to Byar limits |
| `upper_cap`, `root_tol` | `1e8`, `1e-8` | Poisson search cap and inversion tolerance |
| `max_log_odds` | 40 | Absolute finite search limit for binary inversion |

The score floor acts on a **one-sided tail used to form a normal score**, not
on the reported raw two-sided p-value. Theoretical calibration preserves raw
tails. Estimated/fixed calibration uses the finite score and can therefore
limit extreme-tail resolution. Exact zero supplied p-values also need a
finite-score convention. A requested calibrated cutoff beyond the attainable
score range is reported as `SCORE_FLOOR_LIMITS_INVERSION`.

The result is an object of class `fit_pci`:

- `results` contains one row per input provider. Key columns are `z_raw`,
  `p_raw`, `null_mean`, `null_sd`, `z_adjusted`, `p_value`, `direction`, and
  `ci_lower`, `ci_upper`, `ci_scale`. `statistic_value` retains the initial
  test statistic, and `df` records Wald degrees of freedom.
- `included` marks providers with defined scores; `used_for_null` records fit
  selection. With `na_action = "omit"`, incomplete rows and identifiers stay
  present with missing calculated results.
- `null_fit` and `null_parameters` expose the fitted estimator and calibration
  parameters. `as.data.frame(pci)` returns `pci$results`.
- `diagnostics` contains warning codes, omitted identifiers, fit details,
  interval failures, and interpretation notes.
- `method`, `statistic_details`, `call`, and `package_version` record the
  selected methods and relevant inputs/settings.

Inspect `diagnostics$warning_codes` and `diagnostics$intervals` before using the
results. Codes distinguish missing inputs, undefined scores, small-group
fallbacks, nonconvergence, normal-score floors, empty confidence sets, and
failed numerical searches. Undefined or failed intervals remain missing.

Outcome-model uncertainty enters only through the supplied inputs. The
function does not recompute model covariance matrices, refit likelihood-ratio
tests, or identify latent patient-level confounding. Structured null
parameters describe a provider-level working distribution; substantive and
causal interpretation requires additional assumptions outside this interface.
