---
title: "Empirical-null models and diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Empirical-null models and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(EmpiNull)
```

EmpiNull estimates reference distributions for provider-level test statistics.
An empirical null can account for a shared shift, extra spread, or relationships
with provider size and measured provider summaries. The package works after the
primary outcome model has been fitted: it does not fit that model or reconstruct
its standard errors from patient records.

Use `fit_pci()` for tests, null calibration, and supported confidence intervals
in one result. Use `fit_empirical_null()` when your main interest is a structured
null model and its diagnostics. This article explains the calibration models;
the `fit_pci` article describes the testing interface, input scales, and
interval choices.

## What is being estimated?

Let $z_i$ be the statistic for provider $i$. A Gaussian working null specifies

$$Z_i \mid H_{0i} \sim N(\mu_i,\sigma_i^2).$$

For Gaussian calibration, the corrected score and its two-sided p-value are

$$z_i^{\mathrm{adj}}=\frac{z_i-\widehat\mu_i}{\widehat\sigma_i},
\qquad p_i=2\Phi(-|z_i^{\mathrm{adj}}|).$$

With theoretical calibration, `fit_pci()` preserves the selected raw test's
p-value, including discrete or finite-degrees-of-freedom tests. The normal-tail
formula above describes empirical or fixed Gaussian calibration.

The direction depends on the original statistic. Positive scores can represent
higher rates, larger effects, or another quantity defined by the primary model;
they do not automatically mean better performance.

The fit describes a provider-level reference distribution. It does not identify
an unmeasured patient characteristic, prove that confounding has been removed, or
separate a causal quality effect from every other source of variation.
Calibration is useful only when the chosen null structure reasonably describes
the providers being compared.

## Available calibration models

Here $s_i>0$ is `size`, $x_i$ is a row of the provider design matrix, and $g(i)$
identifies a group.

| Model | Null mean $\mu_i$ | Null variance $\sigma_i^2$ | Entry point |
|:--|:--|:--|:--|
| Theoretical | $0$ | $1$ | `fit_pci(calibration = "theoretical")` |
| Fixed | Supplied `null_mean` | Supplied `null_sd` squared | `fit_pci(calibration = "fixed")` |
| Overall | Shared estimated $\mu$ | Shared estimated $\sigma^2$ | Both interfaces |
| Groupwise | Estimated $\mu_{g(i)}$ | Estimated $\sigma_{g(i)}^2$ | `fit_pci(calibration = "groupwise")` |
| Individualized | $0$, or an optional common mean | $1+\phi s_i$ | Both interfaces |
| Correlated | $\sqrt{s_i}\,x_i^\top\zeta$ | $1+\phi s_i$ | Both interfaces |
| CRE | $0$ | $\phi s_i$ | Both interfaces |
| Poisson moment | Poisson-lognormal score mean | Poisson-lognormal score variance | Both interfaces |

`null_sd` values are **standard deviations**. `phi`, `sigmaAlpha2`, and
`sigmaEps2` are **variance components**. For example, the individualized model
uses `sqrt(1 + size * phi)` as its null SD, not `1 + size * phi` and not
`sqrt(1 + size * phi^2)`.

For the correlated model, the default intercept is a column in $x_i$. Its
contribution is $\sqrt{s_i}\zeta_0$, which differs from a shared constant mean.
To estimate a constant mean with individualized variance through `fit_pci()`,
set `control = list(common_intercept = TRUE)`.

`fit_empirical_null(model = "auto")` chooses `"moment"` for
`family = "poisson"`; for `family = "gaussian"`, it chooses `"individualized"`
without covariates and `"correlated"` with covariates. This is an input-based
selection rule, not a test of which model fits best. Specifying the model
explicitly makes an analysis easier to review.

## Robust overall and groupwise calibration

Robust calibration estimates an intercept and residual scale with `MASS::rlm()`.
`psi = "bisquare"` is the default; `psi = "huber"` is also available. The scale
estimate uses MAD. These are Gaussian calibrations with robust parameter
estimation, not distribution-free tests.

This example creates 360 providers, with larger null spread among larger
providers and a small set of shifted providers.

```{r groupwise}
set.seed(2026)
m <- 360L
provider_size <- seq(20, 400, length.out = m)
z <- rnorm(m, mean = 0.15, sd = sqrt(1 + 0.008 * provider_size))
z[seq(12, m, by = 24)] <- z[seq(12, m, by = 24)] + 5

group_fit <- fit_pci(
  z = z, calibration = "groupwise", size = provider_size,
  n_groups = 4, psi = "huber"
)
group_fit$null_parameters
head(group_fit$results[, c("provider_id", "group", "null_mean", "null_sd",
                          "z_adjusted", "p_value", "direction")])
group_fit$diagnostics$warning_codes
```

By default, groups follow size quantiles. If ties produce repeated quantile
boundaries, reduce `n_groups`, use `grouping = "rank"`, or supply `groups`.
Rank grouping can split equal sizes; ties follow input order. It uses blocks of
size `ceiling(m / n_groups)`, so the final block may be smaller and fewer than
`n_groups` groups can result.

Each group needs at least three eligible providers by default. This is a
computational minimum, not evidence of stable estimation with three providers.
`control$min_group_size` can enforce a larger minimum. The default is to fail for
a small group; `control = list(small_group = "theoretical")` explicitly uses
$N(0,1)$ for such a group and records a diagnostic warning.

`fit_mask` selects providers used to fit robust overall or groupwise nulls.
Excluded providers still receive test results. Choose exclusions from the study
design or other defensible information; choosing them to obtain preferred flags
changes the meaning of the analysis.

Each group has its own fitted mean by default. `common_mean = 0` fixes all means
to zero; `common_mean = TRUE` uses the arithmetic mean of all eligible finite
scores. Group-specific scales are retained without refitting. This pooled
arithmetic mean is not itself a robust location estimate.

For one shared robust null, use
`fit_pci(z = z, calibration = "overall")`.

## Structured likelihood models

The overall likelihood, individualized, correlated, CRE, and Poisson moment
models estimate parameters using a central region of the score distribution.
A mixture formulation allows a null proportion $p_0$ and assumes that the
central region is dominated by null or near-null providers. The likelihood uses
scores inside the region and the probability of being outside it; it does not
simply discard outside observations.

`p_grid` supplies candidate null proportions. Its default in the unified
interfaces is `seq(0.5, 0.995, by = 0.005)`. `cutoff` controls the coverage used
to initialize provider-specific central intervals for structured models. The
overall likelihood model instead uses a median/IQR-based interval. Estimates
can be sensitive to the central region, search range, and signal contamination.

The overall model differs between the entry points:

* `fit_pci(calibration = "overall")` uses robust estimation by default.
* `fit_pci(calibration = "overall", control = list(estimator = "likelihood"))`
  uses likelihood estimation.
* `fit_empirical_null(model = "overall")` uses likelihood estimation.

Robust models do not estimate `p0`. In likelihood models, `p0` is a mixture
parameter. It is not a provider-specific probability of being null, a local
false discovery rate, or a multiple-testing adjustment.

### Individualized and covariate-dependent calibration

Use a continuous size model when its mean/variance relationship is supported by
the original score and application. `size` is an effective information measure:
it need not be patient count, and it should not be replaced by another size
measure without checking the model.

Here is a synthetic correlated-null example. The modest grid makes the example
quick; a substantive analysis should check sensitivity to the grid and bounds.

```{r correlated}
set.seed(81)
m <- 300L
size <- runif(m, 20, 180)
case_mix <- cbind(case_mix = rnorm(m))
mu <- sqrt(size) * (0.03 + 0.04 * case_mix[, 1])
z_structured <- rnorm(m, mu, sqrt(1 + 0.012 * size))

null_fit <- fit_empirical_null(
  z = z_structured, size = size, covariates = case_mix,
  family = "gaussian", model = "correlated",
  p_grid = c(0.80, 0.90, 0.95, 0.99), phi_bounds = c(0, 0.1)
)
null_fit$parameters
head(predict(null_fit))
head(flag_empirical_null(null_fit))
audit_empirical_null(null_fit)$diagnostics$warning_codes
```

An intercept is added unless the design already has a constant column or
`include_intercept = FALSE`. Covariates must be numerical, have one row per
provider, and form an identifiable design. Centering/scaling nonconstant
covariates can help conditioning and interpretation.

The same model is available from the testing interface:

```{r correlated-pci}
pci_fit <- fit_pci(
  z = z_structured, calibration = "correlated", size = size,
  covariates = case_mix,
  control = list(p_grid = c(0.80, 0.90, 0.95, 0.99),
                 phi_bounds = c(0, 0.1))
)
head(pci_fit$results[, c("provider_id", "null_mean", "null_sd",
                        "z_adjusted", "p_value")])
```

Without a covariate-dependent mean, use `model = "individualized"` or
`calibration = "individualized"`. The CRE model uses variance $\phi s_i$,
**without the additive 1**. Use it only for a score whose derivation supports
this variance structure. Choosing `"cre"` does not fit a correlated random-effects
outcome model or automatically remove mean bias.

`predict()` returns null parameters for the fitted providers; it does not accept
new covariates for out-of-sample prediction. Replacement `z` values in
`standardize_empirical_null()` or `flag_empirical_null()` must correspond to
the same providers in the same order.

## Poisson-lognormal moment calibration

This model is designed for the Pearson score

$$Z_i=\frac{O_i-E_i}{\sqrt{E_i}}, \qquad E_i=\sum_j r_{ij},$$

where $r_{ij}>0$ are expected contributions from the primary risk model. The
additional input is $E_i^{(2)}=\sum_j r_{ij}^2$. `second_moment` is this sum of
squared contributions, not $E_i^2$, the variance of $E_i$, or a second count
outcome. Preserve it when aggregating primary model output.

For the working model
$\lambda_{ij}=r_{ij}\exp(x_i^\top\zeta+a_i+\varepsilon_{ij})$, let
$a_i\sim N(0,A)$ be a shared provider effect and
$\varepsilon_{ij}\sim N(0,B)$ independent patient effects, independent of $a_i$.
Counts are conditionally independent Poisson given these effects. Define
$M_i=\exp\{x_i^\top\zeta+(A+B)/2\}$. The score moments are

$$\mu_i=\sqrt{E_i}(M_i-1),$$

$$\sigma_i^2=M_i+M_i^2\left[
   (e^A-1)E_i+e^A(e^B-1)\frac{E_i^{(2)}}{E_i}
\right].$$

`sigmaAlpha2` is $A$ and `sigmaEps2` is $B$. These moments are exact under the
specified model, but the Gaussian distribution used for calibration is a
**moment-matched working null**. It is not an exact Poisson test or an exact
Poisson-lognormal distribution.

This reproducible construction retains both required summaries:

```{r poisson-moments}
set.seed(92)
m <- 240L
n <- sample(30:100, m, replace = TRUE)
expected_contributions <- lapply(n, function(ni) {
  exp(rnorm(ni, mean = -0.5, sd = 0.6))
})
E <- vapply(expected_contributions, sum, numeric(1))
E2 <- vapply(expected_contributions, function(x) sum(x^2), numeric(1))
X <- cbind("(Intercept)" = 1, case_mix = rnorm(m))

working_moments <- poisson_null_moments(
  Xbar = X, E = E, E2 = E2, zeta = c(0.02, 0.03),
  sigmaAlpha2 = 0.005, sigmaEps2 = 0.02
)
head(data.frame(mean = working_moments$mean, sd = working_moments$sd))
```

To fit from observed counts, the testing interface constructs a Pearson score.
Here counts are simulated under a simple Poisson null to illustrate the input
contract; the example does not establish recovery of nonzero variance components.

```{r poisson-fit}
O <- rpois(m, E)
moment_fit <- fit_pci(
  observed = O, expected = E, statistic = "poisson_score",
  calibration = "moment", covariates = X, second_moment = E2,
  ci = "none", control = list(p_grid = c(0.80, 0.90, 0.95, 0.99))
)
moment_fit$null_parameters
moment_fit$diagnostics$warning_codes
```

The example may select the largest candidate `p0`, producing
`P0_AT_SEARCH_BOUNDARY`. All counts here were generated under a null model, so
the behavior illustrates why a boundary diagnostic needs interpretation rather
than automatic dismissal of the result.

For this route, `size` defaults to `expected` and the reference ratio must be 1.
Alternatively, pass the Pearson score to `fit_empirical_null()` with
`family = "poisson"`, `model = "moment"`, `size = E`, and
`second_moment = E2`. A mid-p normal score is a different statistic and must not
be substituted into these moment formulas. For supplied `z`, `fit_pci()` requires
`control = list(score_type = "pearson")` to make that requirement explicit.

Separate identification of $A$ and $B$ depends on variation in $E_i$ and
$E_i^{(2)}/E_i$. Strongly related summaries can produce unstable component
estimates even when fitted null means and standard deviations seem plausible.

## Reading diagnostics

Inspect diagnostics before interpreting flags. `fit_empirical_null()` returns
`status`, `diagnostics`, and `low_level_fit`; `audit_empirical_null()` collects
estimates, assumptions, and diagnostics. `fit_pci()` returns `diagnostics` and
retains the fitted model in `null_fit`.

| Diagnostic code | What to investigate |
|:--|:--|
| `OPTIMIZATION_FAILED` / `ROBUST_DID_NOT_CONVERGE` | Convergence, data scale, and starting model assumptions. |
| `P0_AT_SEARCH_BOUNDARY` | Sensitivity to candidate null proportions and the central-region assumption. |
| `PHI_AT_SEARCH_BOUNDARY` | Restrictive variance search bounds or a fit favoring a boundary. |
| `ILL_CONDITIONED_DESIGN` / `FEW_PROVIDERS_PER_PARAMETER` | Redundant covariates, limited information, and overparameterization. |
| `SMALL_CENTRAL_SAMPLE` / `CENTRAL_SKEWNESS` | Whether enough central observations support a roughly symmetric working null. |
| `POISSON_VARIANCE_COMPONENTS_WEAKLY_IDENTIFIED` | Whether expected-count summaries distinguish the two variance components. |

These thresholds are screening rules, not formal model-selection tests. A
warning can occur in a useful fit, and a warning-free fit does not prove correct
calibration. The central mean and SD describe a truncated subset; its SD is not
expected to be 1 even under a correct normal model. Size-subgroup summaries use
all standardized scores and can reflect outliers as well as poor calibration.

Compare reasonable specifications and inspect score distributions by provider
size. Report the statistic, reference, size definition, covariates, calibration
method, tuning/search settings, and diagnostic codes. Tests and intervals treat
fitted null parameters as given; they do not propagate their estimation
uncertainty. Reported p-values are unadjusted for multiple testing.

## Lower-level functions

Original estimators remain exported for analyses that need their specific
controls. Return structures and default grids differ between functions.

| Function | Role |
|:--|:--|
| `empirical_null_overall()` | Overall likelihood null. |
| `empirical_null_groupwise()` | Robust size-stratified null. |
| `empirical_null_ind_RQ()` | Individualized variance, with optional constant mean. |
| `empirical_null_ind_CF()` | Covariate-dependent mean and individualized variance. |
| `empirical_null_ind_CF_CRE()` | Zero-mean CRE variance model. |
| `empirical_null_ind_CF_pois()` | Simplified Poisson calibration using expected counts as size. |
| `empirical_null_ind_pois()` | Poisson-lognormal moment calibration. |
| `empirical_null_CF_prov()` | Specialized count-based estimator with observed provider covariates and coefficient covariance estimates. |

The last specialized estimator is not a separate `fit_pci()` calibration option.
See each function's help for its input contract. The unified interfaces provide
consistent results for supported models without removing lower-level functions.
